Structure-Preserving Model-Reduction of Dissipative Hamiltonian Systems

### 2017-03-27 dissipative_hamiltonian/chapters/3.model.tex
second edit finished

For parametric problems, evaluation of the projection error in (\ref{eq:3.4}) for the entire parameter space is computationally demanding. One can use the loss in the Hamiltonian function as a cheap surrogate to the projection error, i.e.
\begin{equation}
	\omega^{k+1} := \underset{\omega\in \Omega}{\text{argmax }} | H(z(\omega)) - H(AA^+z(\omega)) |.
\end{equation}
Here $\Omega$ is a closed and bounded set of parameters for the original Hamiltonian system. Note that since the Hamiltonian function is constant in time, then $\omega^{k+1}$ can be identified prior to time integration, i.e. $H(z) = H(z_0)$ and $H(AA^+z) = H(AA^+z_0)$. We summarize the greedy method for symplectic basis generation in Algorithm \ref{alg:added1} and refer the reader to \cite{Maboudi:2016} for further details.

@@

The main advantage of the RDH method compared to the existing method is that it enables the reduced system to be integrated using a symplectic integrator. The reduced system constructed using the RDH is a closed Hamiltonian system, therefore the conservation law in Theorem \ref{theorem:2.1} holds and a symplectic integrator guarantees that the total energy is preserved in the reduced system. Alternative methods, e.g. \cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration respectively.

>>>>

For parametric problems, evaluation of the projection error in (\ref{eq:3.4}) for the entire parameter space is computationally demanding. One can use the loss in the Hamiltonian function as a cheap surrogate to the projection error, i.e.
\begin{equation}
	\omega^{k+1} := \underset{\omega\in \Omega}{\text{argmax }} | H(z(\omega)) - H(AA^+z(\omega)) |.
\end{equation}
Here $\Omega$ is a closed and bounded set of parameters for the original Hamiltonian system. Note that $\omega^{k+1}$ can be identified prior to time integration since the Hamiltonian function is constant in time, i.e. $H(z) = H(z_0)$ and $H(AA^+z) = H(AA^+z_0)$. We summarize the greedy method for symplectic basis generation in Algorithm \ref{alg:added1} and refer the reader to \cite{Maboudi:2016} for further details.

@@

The main advantage of the RDH method compared to the existing methods is that it enables the reduced system to be integrated using a symplectic integrator. The reduced system constructed using the RDH is a closed Hamiltonian system, therefore the conservation law in Theorem \ref{theorem:2.1} holds and a symplectic integrator guarantees that the total energy is preserved in the reduced system. Alternative methods, e.g. \cite{peng2016geometric,Polyuga:2010gj,beattie2011structure}, either integrate the reduced system with a non-symplectic integrator, or do not construct a closed reduced system which result in accumulation of local errors or unstable solution during long time integration respectively \cite{Hairer:1250576}.

### 2017-03-27 dissipative_hamiltonian/chapters/4.numer.tex
second edit finished

\subsection{Dissipative wave equation}

@@

The $L^2$-error between the full systems and the RDH, the PSD, and the POD methods are presented in Figure \ref{fig:4.1}.(c). We notice that the symplectic methods provide a more accurate solution when compared to the POD method. In fact, the POD method does not yield a stable reduced system.  Furthermore, it is seen that enriching the PSD reduced basis does not yield a significant improvement in the accuracy of the reduced system. This happens as the PSD method, numerically integrate a non-conservative system with a symplectic integrator. This results in an incorrect evolution of the energy and eventually, in a qualitatively wrong numerical solution.

@@

@@

\begin{figure}[t]
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{./figs/porthamil/error_capac} & 
\includegraphics[width=0.5\textwidth]{./figs/porthamil/error_flux} \\
(a) capacitors & (b) inductors \\
\multicolumn{2} {c} {\includegraphics[width=0.5\textwidth]{./figs/porthamil/error1}} \\
\multicolumn{2} {c} {(c) charge in $C_1$} 
\end{tabular}
\caption{Error between the full model and the reduced model obtained by the reduced dissipating Hamiltonian method ``RDH'' and the moment matching method ``MM''. (a) The average error of charge in capacitors in time. (b) the average error of the flux in inductors in time. (c) Error in $C_1$.} \label{fig:4.3}
\end{figure}

>>>>

\subsection{Dissipative wave equation} \label{sec:4.1}

@@

The $L^2$-error between the full system and the RDH, the PSD, and the POD methods are presented in Figure \ref{fig:4.1}.(c). We notice that the symplectic methods provide a more accurate solution when compared to the POD method. In fact, the POD method does not yield a stable reduced system.  Furthermore, it is seen that enriching the PSD reduced basis does not yield a significant improvement in the accuracy of the reduced system. This happens as the PSD method, numerically integrate a non-conservative system with a symplectic integrator. This results in an incorrect evolution of the energy and eventually, in a qualitatively wrong numerical solution.

@@

\subsection{The sine-Gordon equation}
Consider the one-dimensional dissipative nonlinear wave equation
\begin{equation} \label{eq:added.4.1}
	\left\{
	\begin{aligned}
		q_{t}(t,x) &= p(t,x), \\
		p_{t}(t,x) &= q_{xx}(t,x) - \sin(q) - r(x)  p(t,x), \\
		q(0,x) &= q_0(x), \quad q(t,0) = a, \quad q(t,L) = b,\\
		p(0,x) &= p_0(x).
	\end{aligned}
	\right.
\end{equation}
defined on a domain of length $L$, which is known as the sine-Gordon equation. In the absence of dissipation, $r(x) = 0$, the \emph{kink} solution to (\ref{eq:added.4.1}) is given as
\begin{equation} \label{eq:added.4.2}
	q(t,x) = 4\arctan\left( \exp \left( \frac{(x-x_0 - vt)}{\sqrt{1-v^2}} \right) \right),
\end{equation}
where $|v|< 1$ is the wave speed. In the presence of dissipation, where $r(x)\geq 0$, the traveling wave de-accelerates and stops. The TDD formulation for (\ref{eq:added.4.1}) takes the form
\begin{equation} \label{eq:added.4.3}
	\dot z = \mathbb J_{2n} K^T f(t) + \mathbb J_{2n} g(z) - \mathbb J_{2n} z_{\text{bd}}, \quad f(t) + R \int_0^t f(s) \ ds = K z.	
\end{equation}
where $z$, $K$, $R$ and $f$ are defined similar to (\ref{eq:4.4}), with $r_{\Delta} = rI_{n}$, and 
\begin{equation}
g(z) = (\sin(q_1),\dots,\sin(q_{N_{\Delta x}}),0,\dots,0)^T,
\end{equation}
and $z_{\text{bd}}$ is the term corresponding to the Dirichlet boundary condition. Note that the extended Hamiltonian $H_{ex}$ takes the form
\begin{equation}
	H_\text{ex}(z,\phi,\theta) = \frac 1 2 \left( \| Kz - \phi(t,0) \|_2^2 + \| G(z) \|_2^2 + \| \theta(t) \|^2_{\mathcal H^{2n} } + \| \partial_x\phi(t)\|^2_{\mathcal H^{2n} }\right),
\end{equation}
where $G(z)$ is a potential for $g(z)$ given as $G(q_i) = 1 - \sin(q_i)$, for $i=1,\dots,N_{\Delta_x}$. System parameters are summarized below

\vspace{0.5cm}
\begin{center}
\begin{tabular}{|l|l|}
\hline
Domain length & $L = 50$ \\
No. grid points & $N = 500$ \\
Space discretization size & $\Delta x = L/N$ \\
Time discretization size & $\Delta t = 0.02$ \\
Wave speed & $v = 0.5$ \\
Boundary conditions & $a = 0$, $b=1$\\
Dissipation coefficient & $r = 0.1$ \\
\hline
\end{tabular}
\end{center}
\vspace{0.5cm}

The RDH reduced system is constructed following Algorithm \ref{alg:3.1}. To reduce the complexity of the nonlinear term, we used the symplectic discrete empirical interpolation method (SDEIM) \cite{Maboudi:2016}. The performance of the method is then compared to the PSD and the POD where the SDEIM proposed in \cite{Peng:2014di} and the classical DEIM \cite{Chaturantabut:2010cz} is applied to reduce the complexity of the nonlinear term, respectively.

\begin{figure}[t]
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/error} & 
\includegraphics[width=0.5\textwidth]{./figs/wave_nonlin/energy} \\
(a) & (b)
\end{tabular}
\caption{(a) The $L^2$-error for the different methods, (a) Evolution of error in the kinetic energy for different methods.} \label{fig:added4.1}
\end{figure}

Figure \ref{fig:added4.1}.(a) shows the The $L^2$-error between the full system and the RDH, the PSD, and the POD methods. Although the Hamiltonian system of the sine-Gordon equation is nonlinear, the errors for the different methods show a similar behavior as those in Section (\ref{sec:4.1}). We observe that the POD does not yield a stable reduce system while the symplectic methods provide a high accuracy. Furthermore, we notice that enriching the PSD basis does not significantly enhance the accuracy of the method.

The evolution of error in the kinetic energy $K(p) = \|p\|_2^2/2$ is illustrated in Figure \ref{fig:added4.1}.(b). We see that the POD does not conserve the evolution of the kinetic energy. The RDH method conserves the kinetic energy of the system with higher accuracy than the PSD method. Furthermore, the accuracy of the RDH method is better scaled under enrichment of the reduced basis, compared to the PSD method.

It is observed in Figure 1 that the symplectic treatment of the nonlinear terms is essential in correct model reduction of Hamiltonian systems. In addition, the SDEIM can be combined with the RDH method to construct a reduced Hamiltonian system that can be integrated using a symplectic integrator. Thus, the combination preserves the system energy and the symplectic symmetry of Hamiltonian systems.

@@

\begin{figure}[t]
\begin{tabular}{cc}
\includegraphics[width=0.5\textwidth]{./figs/porthamil/error_capac} & 
\includegraphics[width=0.5\textwidth]{./figs/porthamil/error_flux} \\
(a) capacitors & (b) inductors \\
\multicolumn{2} {c} {\includegraphics[width=0.5\textwidth]{./figs/porthamil/error1}} \\
\multicolumn{2} {c} {(c) charge in $C_1$} 
\end{tabular}
\caption{Error between the full model and the reduced model obtained by the reduced dissipative Hamiltonian method ``RDH'' and the moment matching method ``MM''. (a) The average temporal error of charge in capacitors. (b) the average temporal error of the flux in inductors. (c) The error in $C_1$.} \label{fig:4.3}
\end{figure}

### 2017-03-27 dissipative_hamiltonian/paper.tex
second edit finished

\begin{abstract}
Reduced basis methods are popular for approximately solving large and complex systems of differential equations. However, conventional reduced basis methods do not generally preserve conservation laws and symmetries of the full order model. Here, we present an approach for reduced model construction, that preserves the symplectic symmetry of dissipative Hamiltonian systems. The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath. This results in a correct dissipation of energy, preservation of total energy and, ultimately, in the stability of the solution. Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.

>>>>

\begin{abstract}
Reduced basis methods are popular for approximately solving large and complex systems of differential equations. However, conventional reduced basis methods do not generally preserve conservation laws and symmetries of the full order model. Here, we present an approach for reduced model construction, that preserves the symplectic symmetry of dissipative Hamiltonian systems. The method constructs a closed reduced Hamiltonian system by coupling the full model with a canonical heat bath. This allows the reduced system to be integrated with a symplectic integrator, resulting in a correct dissipation of energy, preservation of total energy and, ultimately, in the stability of the solution. Accuracy and stability of the method are illustrated through the numerical simulation of the dissipative wave equation and a port-Hamiltonian model of an electric circuit.
